Visual diagnostics for constrained optimisation with application to guided tours

### 2020-12-10 paper/paper.Rmd
update of optimisation section

Optimization problems are ubiquitous in many areas of study. While in some cases, analytical solutions can be found, the majority of problems rely on numerical methods to find the optimal solution. These numerical methods follow interactive approaches that aim at finding the optimum by progressively improving the current solution until a desirable accuracy is achieved. Although that principle seems uncomplicated, optimization methods need to deal with many challenges such as the existence of multiple maxima (local and global), the trade off between desirable accuracy and computational burden, as well as dealing with noisy objective functions and possible constraints. In addition, optimization results might depend on the algorithm starting values affecting the consistency of results.

Our interest is on constrained optimization [@bertsekas2014constrained] as defined in the introduction section and assume that it is not possible to find a solution to the problem in the way of a closed form. That is, the problem consists of finding the minimum or maximum of a function $f \in L^p$ in the constrained $\mathbb{A}$ space.

@@

Optimization methods can be divided in various classes, such as global optimisation [@kelley1999iterative; @fletcher2013practical], convex optimisation [@boyd2004convex], stochastic optimisation [@nocedal2006numerical] . In this paper, we focused on methods that rely on the objective function derivatives known as gradient methods, and those which are known as derivative free methods where optimization does not rely on the gradient or derivatives information.

In gradient ascent methods the goal is to find the maximum of an objective function while in gradient descent methods the interest is in the minimum. The maximum is searched for by taken steps  proportional to the gradient of the function at the evaluation point that is $x_{k+1} = x_k +\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate [@fletcher2013practical].  The step size is the key in the success of the optimization problem as inadequate step sizes might lead to a failure in convergence. These methods are known to be relative slow and can have convergence issues when problems are poorly conditioned [@trefethen1997numerical]. In order to solve some of those problems, line minimization or line search methods were proposed [@shi2004convergence] where the step size is chosen in such a way that convergence is ensured. In line search methods users are required to provide an initial estimate $x_{1}$, a search direction $S_k$ for each iteration where the objective is to find the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then move on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and repeat the process until the desire convergence is reached.
Modern development of line search methods focuses on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational storing
burden of the intermediate computations while catering for practical optimisation problems. In the next section two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

Although gradient optimization methods are very popular, they rely on the existence of the objective function derivatives and on the complexity of the constraints. Because of that, the second big class of optimization problems look at derivative free methods where the emphasis is on finding, in most cases, a near optimal solution. Examples of those include response surface methodology [@box1951experimental], stochastic approximation [@robbins1951stochastic], random search [@fu2015handbook] and  heuristic methods [@sorensen2013metaheuristics]. Later, we will present an optimization algorithm within the random search class, namely simulated annealing for optimization with the guided tour. Moreover, we will introduce an algorithm to further improve optimization solutions both in gradient descent methods and in derive free optimization.

Several R implementations address optimization problems with both general purpose as well as task specifics solvers. The most prominent one within the general solvers is `optim()` in the \CRANpkg{stats} [@stats] package, which have functions for both gradient based and derivative free methods. Another general solver specialised in non-linear optimisation is \CRANpkg{nloptr} [@nloptr]. For more specific optimisation solvers, readers are recommended to svisite the relevant sections in CRAN task review on *optimisation and mathematical programming* [@crantaskreviewoptim].

>>>>

Optimization problems are ubiquitous in many areas of study. While in some cases analytical solutions can be found, the majority of problems rely on numerical methods to find the optimal solution. These numerical methods follow iterative approaches that aim at finding the optimum by progressively improving the current solution until a desirable accuracy is achieved. Although that principle seems uncomplicated, optimization methods need to deal with many challenges such as the possible existence of multiple maxima (local and global), the trade-off between desirable accuracy and computational burden, as well as dealing with noisy objective functions and possible constraints. In addition, the optimization results might depend on the algorithm starting values, affecting the consistency of results.

Our interest is on constrained optimization [@bertsekas2014constrained] as defined in the introduction section, and assuming it is not possible to find a solution to the problem in the way of a closed-form. That is, the problem consists of finding the minimum or maximum of a function $f \in L^p$ in the constrained $\mathbb{A}$ space.

@@

Optimization methods can be divided into various classes, such as global optimisation [@kelley1999iterative; @fletcher2013practical], convex optimisation [@boyd2004convex] or stochastic optimisation [@nocedal2006numerical]. In this paper, we consider methods that rely on the objective function derivatives known as gradient methods, as well as on methods known as derivative-free methods where optimization does not rely on the knowledge of the gradient or other derivatives.

In gradient ascent methods the goal is to find the maximum of an objective function while in gradient descent methods the interest is in the minimum. The maximum is searched for by taking steps proportional to the gradient of the function at the evaluation point, that is $x_{k+1} = x_k +\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate [@fletcher2013practical]. One difficulty is the choice of step size, as inadequate step sizes might prevent the optimisation method to converge to an extremum. Moreover, these methods are known to be relative slow and can have convergence issues when problems are poorly conditioned [@trefethen1997numerical]. In order to solve some of those problems, line minimization or line search methods were proposed [@shi2004convergence] where the step size is chosen in such a way that convergence is ensured. In line search methods, users are required to provide an initial estimate $x_{1}$ and a search direction $S_k$ for each iteration. Then, the objective is to find the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then one moves on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and the process is repeated until the desire convergence is reached.
Modern developments of line search methods focus on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational burden of storing intermediate computations while catering for practical optimisation problems. In the next section, two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

Although gradient optimization methods are very popular, they rely on the availability of the objective function derivatives and on the complexity of the constraints. Because of that, a large class of derivative-free optimization methods have been developed over the years, where the emphasis is on finding, in most cases, a near optimal solution. Examples of those include response surface methodology [@box1951experimental], stochastic approximation [@robbins1951stochastic], random search [@fu2015handbook] and  heuristic methods [@sorensen2013metaheuristics]. Later, we will present a simulated annealing optimisation algorithm, which belongs to the class of random search methods, for optimisation with the guided tour. Moreover, we will introduce an algorithm to further improve optimization solutions with both gradient descent methods and derivative-free optimisation.

Several R implementations address optimization problems with both general purpose as well as task specifics solvers. The most prominent one within the general solvers is `optim()` in the \CRANpkg{stats} [@stats] package, which provides both gradient-based and derivative-free optimisation functions. Another general solver specialised in non-linear optimisation is \CRANpkg{nloptr} [@nloptr]. For more specific optimisation solvers, readers are recommended to visit the relevant sections in CRAN task review on *optimisation and mathematical programming* [@crantaskreviewoptim].
